---
title: "Scoring the HiTOP-SR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the HiTOP-SR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r}
#| include: false
library(tibble)
```


The HiTOP-SR instrument has 405 items and yields 76 scale scores. To demonstrate the ability of the package to calculate these scale scores, we can use real example data (n=411) that was collected at the University of Kansas (KU) by Girard & Gray in 2024--2025. This data is stored in the package under the name `ku_hitopsr`.

First, we load the package into memory using the `library()` function. If this doesn't work, make sure you installed the package properly (see the README on [GitHub](https://github.com/jmgirard/hitop)).

```{r}
library(hitop)
```

Next, we can load the example dataset from the package using the `data()` function. It is a large tibble that contains a `participant` column with a unique identifier for each participant, a `biosex` column indicating whether each participant is "female" or "male", and then 405 columns numbered `hsr001` to `hsr405` containing each participant's rating on each item of the HiTOP-SR (on a numerical scale from 1 to 4).

```{r}
data("ku_hitopsr")
ku_hitopsr
```

## Basic Scoring

To turn these item-level ratings into mean scores on the 76 scales, we can use the `score_hitopsr()` function. It needs to know what object contains the data and which columns contain the item-level data. There are several ways we can specify the items. First, we can provide the column numbers and use the `:` shortcut. In this tibble, the items are from column 3 to column 407 so we can use `items = 3:407`. I am going to also set `append = FALSE` so that you can quickly see the scale scores.

```{r}
scores <- score_hitopsr(
  data = ku_hitopsr,
  items = 3:407,
  append = FALSE
)
scores
```

## Appending

If I had instead set `append = TRUE` (or left it off, as that is the default), we would get back the `ku_hitopsr` tibble with the scale scores added to the end as extra columns. Notice below how we now have 483 columns instead of 407.

```{r}
scores <- score_hitopsr(
  data = ku_hitopsr,
  items = 3:407
)
scores
```

## Items as Strings

Alternatively, we could provide the item column names as a character string. Typing out all 405 item names would be a hassle, but luckily this dataset named them consistently so we can build the names automatically using `sprintf()`. If we use the "hsr%03d" format and apply that across the numbers 1 to 405, that will create the zero-padded column names we need. If there was no zero-padding, we could have just used "hsr%d".

```{r}
scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405),
  append = FALSE
)
scores
```

## Scale Prefixes

Also note that each scale column has the prefix "hsr_" in its name. You can change the prefix (e.g., setting it to `"hitop_"`) or even turn it off (e.g., setting it to `""`) using the `prefix` argument.

```{r}
scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405),
  prefix = "hitop_",
  append = FALSE
)
scores
```

## Simple Standard Errors

In addition to calculating each scale score as the mean of its corresponding items, we can also calculate a standard error for each scale score: the SD of the items the respondent actually answered divided by the square root of how many of those items they answered. We turn this on using `calc_se`.

Note what these numbers do and do not describe. Each one summarizes how much a respondent's answers varied within a scale; it is not an estimate of how precisely the scale measures the underlying trait, so it does not give a confidence interval for a respondent's true score. For measurement precision, see the reliability coefficients below.

```{r}
scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405),
  calc_se = TRUE,
  append = FALSE
)
scores
```

Note how there are now 152 columns instead of 76. The extra columns aren't shown in the preview above, but they are named with the `_se` suffix, e.g., `hsr_agoraphobia_se`.

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_hitopsr()` function, which returns a tibble with one row per scale and columns for the number of items and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it).

```{r reliability}
#| warning: false
reliability_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405)
)
```

## Scoring a Short Form

Not every study administers all 405 items. The `hitop_subset()` function describes a short form built from a chosen set of scales, and the `generate_docx_hitopsr()`, `generate_qualtrics_hitopsr()`, and `generate_redcap_hitopsr()` functions turn that description into an instrument to field. Data collected that way has only the subset's item columns, so passing it to `score_hitopsr()` as if it were a full administration would fail on the item count.

Instead, hand the same description back through the `subset` argument. Here is a four-scale short form:

```{r}
short_form <- hitop_subset(
  instrument = "hitopsr",
  scales = c(
    "Agoraphobia", "Appetite Loss",
    "Antisocial Behavior", "Romantic Disinterest"
  )
)
short_form
```

We will stand in for the collected data by keeping just those columns of `ku_hitopsr`. Note that the item numbers are the *original* HiTOP-SR numbers — the short form does not renumber its items from 1.

```{r}
short_data <- ku_hitopsr[sprintf("hsr%03d", short_form$items)]
ncol(short_data)
```

Now score it. The `items` argument names the columns you actually have, in instrument order, and `subset` tells the function which scales they belong to.

```{r}
short_scores <- score_hitopsr(
  data = short_data,
  items = names(short_data),
  subset = short_form,
  append = FALSE
)
short_scores
```

Only the subset's scales come back, in the order they appear in `hitopsr_scales`. The values are exactly what a full administration would have produced for those scales — a scale score depends only on its own items, so dropping the other 72 scales' columns cannot move it:

```{r}
full_scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405),
  append = FALSE
)
all.equal(short_scores, full_scores[names(short_scores)])
```

`reliability_hitopsr()` takes the same argument and returns one row per subset scale.

```{r}
#| warning: false
reliability_hitopsr(
  data = short_data,
  items = names(short_data),
  subset = short_form,
  omega = FALSE
)
```

The `srange`, `prefix`, `missing`, `calc_se`, and `append` arguments all behave exactly as they do for a full administration.

## Renaming Item Columns
If your dataset was collected before the item numbers were standardized or used arbitrary variable names, you must rename your columns to the standard format before running the scoring engine. The `rename_hitopsr_items()` function handles this preparation via two approaches.

### Method 1: Legacy "Original" Pool Names
If your columns are labeled with older pool names (e.g., HiTOP_659 or Ext_432), use `method = "original"`. The function scans your data frame columns, matches them against the package's built-in item database, and renames them to standard names.

```{r}
# Setup a mock dataset with older legacy column names
legacy_data <- data.frame(
  HiTOP_659 = c(1, 2, 4),
  HiTOP_301 = c(3, 4, 2),
  ParticipantID = c(101, 102, 103)
)

# Rename legacy columns using the native pipe
standardized_legacy <-
  legacy_data |>
  rename_hitopsr_items(method = "original", prefix = "HSR_")

standardized_legacy
```

### Method 2: Matching via Literal Item Text
If your dataset uses completely customized column names but you tracked the exact question prompts presented to your participants, use `method = "text"`. Provide a vector of your current column names alongside a matching vector of literal item text strings.

```{r}
# Setup a mock dataset with completely custom column names
custom_data <- data.frame(
  q_party = c(1, 2, 1),
  q_flawless = c(4, 3, 4),
  Age = c(21, 25, 30)
)

# Define the mapping pairs
my_cols <- c("q_party", "q_flawless")
my_texts <- c(
  "I preferred to stay home than to go to a party.",
  "I felt that my work must be flawless."
)

# Rename custom columns based on text matching
standardized_custom <- custom_data |>
  rename_hitopsr_items(
    method = "text",
    item_cols = my_cols,
    item_text = my_texts,
    prefix = "HSR_"
  )

standardized_custom
```

*Note: Because these small mock examples contain only a subset of the full item pool, `rename_hitopsr_items()` will safely issue a cli warning letting you know that fewer than 405 items were matched. This perfectly accommodates researchers intentionally administering short forms or separate diagnostic modules.*
