---
title: "Scoring the PID-5-SF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5-SF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hitop)
```

### Score simulated PID-5-SF data

The PID-5-SF is a shorter version of the PID-5 with 100 items that still yields all domain and facet scores. The validity scales are still calculable but may have fewer items and their psychometric properties have not, to my knowledge, been examined with the FSF.

```{r x3}
data("sim_pid5sf")

score_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)

validity_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)
```

### Score real PID-5-SF data

We can repeat this process with real data that was collected at University of Kansas (KU). There should be fewer (but still some) validity problems since this is real data. We can also retain un-scored "ID" variables in the dataset.

```{r x4}
data("ku_pid5sf")

score_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)

validity_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)
```

### Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We can just add one or more of the following arguments to `score_pid5()`: `alpha` and `omega`. For the latter, we will need the **lavaan** package installed. If requested, a table of reliability results will be printed as a side-effect of the function (alongside any warnings from lavaan about convergence of the factor analysis models that omega is based on).

```{r reliability}
#| warning: false
scores <- score_pid5(
  data = ku_pid5sf,
  items = sprintf("pid_%d", 1:100),
  version = "SF",
  alpha = TRUE,
  omega = TRUE
)
```
