---
title: "Scoring the PID-5-SF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5-SF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hitop)
```

```{r}
#| include: false
library(tibble)
```

## Score simulated PID-5-SF data

The PID-5-SF is a shorter version of the PID-5 with 100 items that still yields all domain and facet scores. The validity scales are still calculable but may have fewer items and their psychometric properties have not, to my knowledge, been examined with the FSF.

```{r x3}
data("sim_pid5sf")

score_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)

validity_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)
```

## Score real PID-5-SF data

We can repeat this process with real data that was collected at University of Kansas (KU). There should be fewer (but still some) validity problems since this is real data. We can also retain un-scored "ID" variables in the dataset.

```{r x4}
data("ku_pid5sf")

score_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)

validity_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)
```

## Simple Standard Errors

In addition to calculating each scale score as the mean of its corresponding items, we can also calculate each scale score's standard error as the SD of its corresponding items divided by the square root of its number of items. These standard errors are especially useful when plotting the scores as they can be converted into confidence intervals. We turn this on using `calc_se`.

```{r se}
score_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  calc_se = TRUE,
  append = FALSE
)
```

Note how there are now 60 columns instead of 30. The extra columns aren't all shown in the preview above, but they are named with the `_se` suffix, e.g., `pid_anhedonia_se`.

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_pid5()` function, which returns a tibble with one row per scale and columns for the number of items and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it).

```{r reliability}
#| warning: false
reliability_pid5(
  data = ku_pid5sf,
  items = sprintf("pid_%d", 1:100),
  version = "SF"
)
```
