---
title: "Scoring the PID-5-BF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5-BF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hitop)
```

```{r}
#| include: false
library(tibble)
```

The PID-5-BF (Brief Form) is a 25-item version of the PID-5 that yields the 5 personality-trait domain scores only (it does not produce the 25 facet scores). We can demonstrate the package's functionality using some simulated data.

## Score simulated PID-5-BF data

The `sim_pid5bf` dataset is built into the package and contains 100 rows (each a simulated participant) across 25 columns named `pid_1` to `pid_25`. To compute the 5 domain scores, we use `score_pid5()` with `version = "BF"`. As with the other forms, we can specify the items by column number (`items = 1:25`) and set `append = FALSE` to see just the scores. The only validity scale calculable from this subset of items is the percentage of missing items (PNA), which `validity_pid5()` returns.

```{r}
data("sim_pid5bf")

score_pid5(sim_pid5bf, items = 1:25, version = "BF", append = FALSE)

validity_pid5(sim_pid5bf, items = 1:25, version = "BF", append = FALSE)
```

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_pid5()` function, which returns a tibble with one row per scale and columns for the number of items and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it). Note that, because this is naively simulated data, we would expect the reliability in this example to be poor.

```{r reliability}
#| warning: false
reliability_pid5(
  data = sim_pid5bf,
  items = 1:25,
  version = "BF"
)
```
