---
title: "Scoring the PID-5-SF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5-SF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hitop)
```

```{r}
#| include: false
library(tibble)
```

## Score simulated PID-5-SF data

The PID-5-SF is a shorter version of the PID-5 with 100 items that still yields all domain and facet scores. The validity scales are still calculable but may have fewer items and their psychometric properties have not, to my knowledge, been examined with the FSF.

```{r x3}
data("sim_pid5sf")

score_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)

validity_pid5(sim_pid5sf, items = 1:100, version = "SF", append = FALSE)
```

## Score real PID-5-SF data

We can repeat this process with real data that was collected at University of Kansas (KU). There should be fewer (but still some) validity problems since this is real data. We can also retain un-scored "ID" variables in the dataset.

```{r x4}
data("ku_pid5sf")

score_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)

validity_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  append = FALSE
)
```

## Simple Standard Errors

In addition to calculating each scale score as the mean of its corresponding items, we can also calculate each scale score's standard error as the SD of its corresponding items divided by the square root of its number of items. These standard errors are especially useful when plotting the scores as they can be converted into confidence intervals. We turn this on using `calc_se`.

```{r se}
score_pid5(
  ku_pid5sf,
  items = paste0("pid_", 1:100),
  version = "SF",
  calc_se = TRUE,
  append = FALSE
)
```

Note how there are now 60 columns instead of 30. The extra columns aren't all shown in the preview above, but they are named with the `_se` suffix, e.g., `pid_anhedonia_se`.

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_pid5()` function, which returns a tibble with one row per scale and columns for the number of items and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it).

```{r reliability}
#| warning: false
reliability_pid5(
  data = ku_pid5sf,
  items = sprintf("pid_%d", 1:100),
  version = "SF"
)
```

## Normative Scores

The `pid_norms` dataset carries the normative score distributions published by Markon et al. (2024), including a set built on the short form. The `norm_pid5()` function looks scored columns up in those tables and returns, for each one, the T score and percentile printed against the nearest tabled raw score. It converts scores rather than computing them, so we hand it the output of `score_pid5()` — and of `validity_pid5()`, if we want those scales converted too.

```{r x6}
#| message: false
#| warning: false
scored <- score_pid5(ku_pid5sf, items = paste0("pid_", 1:100), version = "SF")
scored <- validity_pid5(scored, items = paste0("pid_", 1:100), version = "SF")
```

For the short form the published tables cover the five domain scales, all 25 facet scales, and the INC-S inconsistency scale. Each converted scale gains a `_ptl` column, and those whose tables print T scores also gain a `_t` column. INC-S is distributed as percentiles only, so it gets no `_t` column. Note that `version = "SF"` selects the short-form tables — the same raw score converts differently across forms.

```{r x7}
norm_pid5(
  scored,
  scores = paste0(
    "pid_",
    c("negativeAffectivity", "detachment", "antagonism", "disinhibition",
      "psychoticism", "INCS")
  ),
  version = "SF",
  append = FALSE
)
```

Every number returned is a cell of a published table: the nearest printed row is selected and nothing is interpolated. A score that falls outside a printed range is capped to the nearest end rather than extrapolated, and a warning reports how many observations that happened to. A scale the tables do not cover returns `NA` in both columns with a warning naming it. Every report this function makes is a warning, so a single `suppressWarnings()` call silences it.

If the items were answered on a four-option response scale that starts somewhere other than 0 — 1 to 4, say — pass that range as `srange` and each score is reconciled to the published 0–3 metric before it is looked up, with a warning naming which scales were adjusted and which were left alone. The per-scale formulas are given in `?norm_pid5`.
