---
title: "Scoring the PID-5"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| include: false
library(tibble)
```

The Personality Inventory for DSM-5 (PID-5) instrument has 220 items and yields 25 facet scales, 5 domain scales, and 5 validity scales. We can demonstrate the package's functionality using some simulated data.

First, we load the package into memory using the `library()` function. If this doesn't work, make sure you installed the package properly (see the README on [GitHub](https://github.com/jmgirard/hitop)).

```{r}
library(hitop)
```

## Score simulated PID-5 data

The `sim_pid5` dataset is built into the package and can be loaded using the `data()` function. It contains 100 rows (each representing a simulated participant) and 220 columns named `pid5_001` to `pid5_220` (each representing an item from the PID-5).

```{r}
data("sim_pid5")
sim_pid5
```

If your own PID-5 columns are named some other way, `rename_pid5_items()` will rename them to this pattern. By default it reads the item number out of a column already named `pid_1`, `pid_2` and so on -- the spelling this package's own datasets used before they were renamed to match the exports. Setting `method = "text"` instead matches the literal item prompts. Columns it cannot match keep their names and are reported, and a partial rename warns that fewer than 220 items were matched, which suits a study that administered only some of them.

```{r}
old_names <- data.frame(pid_1 = c(0, 1), pid_2 = c(2, 3), age = c(30, 40))

rename_pid5_items(old_names, version = "FULL")
```

`label_pid5()` is the companion helper: it attaches each item's questionnaire prompt to that item's column as a `label` attribute, which data viewers and reporting packages can display in place of the column name. With `target = "scales"` it does the same for the columns `score_pid5()` writes, attaching each facet and domain its display name.

```{r}
labeled <- label_pid5(sim_pid5, target = "items", version = "FULL")

attr(labeled$pid5_001, "label")
```

To turn these item-level data into scale scores on the 25 facets and 5 domains, we can use the `score_pid5()` function. We will need to tell the function which columns contain our items and which version of the PID this is. There are several ways we can specify the items. First, we can provide the column numbers and use the `:` shortcut. In this tibble, the items are from column 1 to column 220 so we can use `items = 1:220`. I am going to also set `append = FALSE` so that you can quickly see the scale scores. I also can set the version to `"FULL"` (or leave that argument off, as that is the default, shown in the example below) to let it know we are using the full 220-item version.

```{r}
scores <- score_pid5(sim_pid5, items = 1:220, version = "FULL", append = FALSE)
scores
```

If I had instead set `append = TRUE` (or left it off, as that is the default), we would get back the `sim_pid5` tibble with the scale scores added to the end as extra columns. Notice below how we now have 250 columns instead of 220 or 30.

```{r}
scores <- score_pid5(sim_pid5, items = 1:220)
scores
```

Alternatively, we could provide the item column names as a character string. Typing out all 220 item names would be a hassle, but luckily this data named them consistently so we can build the names automatically using `sprintf()`. If we use the "pid5_%03d" format and apply that across the numbers 1 to 220, that will create the column names we need.

```{r}
scores <- score_pid5(sim_pid5, items = sprintf("pid5_%03d", 1:220))
scores
```

There are other useful arguments to the function that you can read about using its documentation by typing the following into your R console: `?score_pid5` or through the [package website](https://jmgirard.github.io/hitop/reference/score_pid5.html).

## Rank Each Person's Highest Facets

Thirty scale columns is more than can be read at a glance. The `rank_scales()` function condenses them: for each row it sorts the columns you select and returns a single string naming the highest-scoring ones, highest first. Here we rank the 25 facet scales — the first 25 columns of the scored tibble, the last 5 being the domains those facets roll up into — and ask for each participant's top 5. Setting `prefix = "pid_"` strips that leading string from the column names, so the string reads as scale names.

```{r rank}
facet_scores <- score_pid5(sim_pid5, items = 1:220, append = FALSE)
rank_scales(facet_scores, scales = 1:25, prefix = "pid_", top = 5, append = FALSE)
```

The `top_scales` column holds one comma-separated string per participant: the names of their five highest-scoring facets, in descending order. Ties are broken by the order of the columns you selected. We passed `append = FALSE` so that the ranked column comes back on its own; leaving that argument at its default returns the scored tibble with the column added to the end instead. Use `name` to call the output column something else and `dir = "low"` to rank from the bottom.

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_pid5()` function, which returns a tibble with one row per scale: its printed name (`Scale`), the stem that names its column in the scored output (`camelCase`), the number of items (`nItems`), and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it). Note that, because this is naively simulated data, we would expect the reliability in this example to be poor.

```{r reliability}
#| warning: false
reliability_pid5(
  data = sim_pid5,
  items = sprintf("pid5_%03d", 1:220),
  version = "FULL"
)
```

## Validity Scales for the PID-5

There are also several validity scales that have been developed for the full PID-5, including measures of overreporting, inconsistent responding, and positive impression management. We can use the simulated data to demonstrate the ability of the `validity_pid5()` function to calculate these scores and flag issues. The function arguments will be consistent with what we just learned. Note that, because the data is fake, we would expect there to be lots of validity issues.

```{r x2b}
validity_pid5(sim_pid5, items = 1:220, append = FALSE)
```

## Normative Scores

The `pid_norms` dataset carries the normative score distributions published by Markon et al. (2024). The `norm_pid5()` function looks scored columns up in those tables and returns, for each one, the T score and percentile printed against the nearest tabled raw score. It converts scores rather than computing them, so we hand it the output of `score_pid5()` — and of `validity_pid5()`, if we want those scales converted too.

```{r x2c}
#| message: false
#| warning: false
scored <- score_pid5(sim_pid5, items = 1:220)
scored <- validity_pid5(scored, items = 1:220)
```

For the full form the published tables cover the five domain scales, all 25 facet scales, and three of the validity scales — INC, ORS, and PRD. They do not cover SD-TD at all. Each converted scale gains a `_ptl` column, and those whose tables print T scores also gain a `_t` column. The validity scales are distributed as percentiles only, so they get no `_t` column.

```{r x2d}
norm_pid5(
  scored,
  scores = paste0(
    "pid_",
    c("negativeAffectivity", "detachment", "antagonism", "disinhibition",
      "psychoticism", "INC", "ORS", "PRD")
  ),
  version = "FULL",
  append = FALSE
)
```

The facets convert the same way. Anhedonia, for instance, is an item mean over its eight full-form items, and its table runs from a floor of 0.00 up to a printed 3.84 — past the 3.00 an item mean can actually reach, which is true of most of the facet columns and is a property of the published tables rather than of the conversion:

```{r x2d2}
norm_pid5(scored, scores = "pid_anhedonia", version = "FULL", append = FALSE)
```

Every number returned is a cell of a published table: the nearest printed row is selected and nothing is interpolated. A score that falls outside a printed range is capped to the nearest end rather than extrapolated, and a warning reports how many observations that happened to — the `PRD` sum reaches 66 while its table stops at 55, so it is a common one to see. A scale the tables do not cover returns `NA` in both columns with a warning naming it. Every report this function makes is a warning, so a single `suppressWarnings()` call silences it.

If the items were answered on a four-option response scale that starts somewhere other than 0 — 1 to 4, say — pass that range as `srange` and each score is reconciled to the published 0–3 metric before it is looked up, with a warning naming which scales were adjusted and which were left alone. The per-scale formulas are given in `?norm_pid5`.

## Profile Plots

Once a respondent's scores are normed, `plot_pid5()` draws them as a profile against the published metric. It takes one respondent — a profile plot shows one person — so we norm the whole dataset and hand it a single row.

```{r x2e}
#| message: false
#| warning: false
domains <- paste0(
  "pid_",
  c("negativeAffectivity", "detachment", "antagonism", "disinhibition",
    "psychoticism")
)
normed <- norm_pid5(scored, scores = domains, version = "FULL")
```

```{r x2f}
#| fig.width: 7
#| fig.height: 4
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
plot_pid5(normed[1, ], version = "FULL")
```

The dashed line marks T = 50, the normative sample's mean, and the score axis spans the range the published tables actually print for these scales — so the axis does not rescale from respondent to respondent and two profiles are directly comparable. Nothing on the plot says whether a score is high, low, or concerning: {hitop} presents scores against norms and leaves the interpreting to you.

Passing `level = "facet"` plots all 25 facets instead, grouped into a panel per domain. The APA key ties three facets to each domain; the remaining ten define no domain and are grouped separately rather than dropped.

```{r x2g}
#| fig.width: 7.5
#| fig.height: 9
#| message: false
#| warning: false
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
facets <- paste0("pid_", pid_scales[["FULL"]]$camelCase)
normed_facets <- norm_pid5(scored, scores = facets, version = "FULL")
plot_pid5(normed_facets[1, ], version = "FULL", level = "facet")
```

Set `metric = "percentile"` for a percentile axis instead of T scores. `norm_pid5()` returns percentiles as a proportion; the plot multiplies them by 100 so the axis reads 0–100.

```{r x2h}
#| fig.width: 7
#| fig.height: 4
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
plot_pid5(normed[1, ], version = "FULL", metric = "percentile")
```

The result is an ordinary ggplot object, so you can restyle it with any ggplot2 layer — `+ ggplot2::labs(title = ...)`, a different theme, and so on.

## The PID5BF+M

The PID5BF+M (Bach et al., 2020, [doi:10.1159/000507589](https://doi.org/10.1159/000507589)) is a 36-item form with 18 facets of 2 items each and 6 domains: the 5 PID-5 trait domains and Anankastia. Every item is a PID-5 item, and no item is reverse-keyed. The `pid_items$BFPM` column gives each item's number on this form. The package names its item columns `pid5bfpm_01` to `pid5bfpm_36`.

The package has no simulated BF+M dataset. To have something to score, we take the 36 BF+M items from `sim_pid5` in BF+M order and rename them.

```{r bfpm-data}
bfpm_rows <- pid_items[!is.na(pid_items$BFPM), ]
bfpm_rows <- bfpm_rows[order(bfpm_rows$BFPM), ]
sim_bfpm <- sim_pid5[sprintf("pid5_%03d", bfpm_rows$FULL)]
names(sim_bfpm) <- sprintf("pid5bfpm_%02d", bfpm_rows$BFPM)
```

`score_pid5()` with `version = "BFPM"` returns the 18 facets, then the 6 domains, in the order of the form's key. Each facet is the mean of its 2 items, and each domain is the mean of its 3 facets. The map from domains to facets is in `pid_bfpm_domains`.

```{r bfpm-score}
bfpm_scores <- score_pid5(sim_bfpm, items = 1:36, version = "BFPM", append = FALSE)
bfpm_scores
```

These are item means on the 0 to 3 scale, as for the other forms. The form's published key sums the 2 items of a facet and averages the facet sums for a domain. On complete data, its values are twice the package's:

```{r bfpm-sums}
head(2 * bfpm_scores$pid_anankastia)
```

No missing-data rule is published for the PID5BF+M. Under the default `missing = "apa"`, any missing item makes its facet `NA`, and an `NA` facet makes its domain `NA`. With whole-number responses that is the same output as `missing = "complete"`. With `missing = "available"`, a facet can come from one item and a domain from one or two facets.

`reliability_pid5()` returns the 18 facets and then the 6 domains. Omega is `NA` for every 2-item facet, because a one-factor model of 2 items is not identified. These simulated responses are random, so the estimates here are poor: some domain omegas also fail and come back `NA`, and the chunk hides the warnings these fits raise.

```{r bfpm-reliability}
#| warning: false
print(reliability_pid5(sim_bfpm, items = 1:36, version = "BFPM"), n = 24)
```

`rename_pid5_items()` and `label_pid5()` also take `version = "BFPM"`. The PID5BF+M has no validity scales, and the package has no norms or profile plot for it.

## The PID-5 Informant Form

The PID-5 Informant Form (PID-5-IRF; Markon et al., 2013, [doi:10.1177/1073191113486513](https://doi.org/10.1177/1073191113486513)) is the APA's 218-item form on which an adult informant rates the person receiving care. It has the full form's 25 facets and 5 domains. It has no counterpart to self-report items 96 and 177, so its Anxiousness and Suspiciousness facets each have one item fewer. The `pid_items$IRF` column gives each informant item's number, `pid_items$TextIRF` its wording (each item completes the stem "He or she..."), and the package names its item columns `pid5irf_001` to `pid5irf_218`.

The package has no simulated informant dataset. For illustration, we take the 218 self-report items that the informant items map to from `sim_pid5` and rename them.

```{r irf-data}
irf_rows <- pid_items[!is.na(pid_items$IRF), ]
sim_irf <- sim_pid5[sprintf("pid5_%03d", irf_rows$FULL)]
names(sim_irf) <- sprintf("pid5irf_%03d", irf_rows$IRF)
```

`score_pid5()` with `version = "IRF"` returns the same 25 facets and 5 domains as the full form, under the same column names, so informant and self-report scores of the same person line up column by column.

```{r irf-score}
irf_scores <- score_pid5(sim_irf, items = 1:218, version = "IRF", append = FALSE)
irf_scores
```

Scoring follows the APA informant key (Markon et al., 2013), with two readings that `?score_pid5` explains. Fourteen items are reverse-scored, the ones the key's Facet Table marks R; the key's Step 1 list also names items 98 and 176, which are not reversed. And the key says to "round up" a fractional prorated raw score; the package applies the nearest-whole-number rule of the other APA PID-5 keys, so informant and self-report facets prorate the same way.

`reliability_pid5()`, `rename_pid5_items()` and `label_pid5()` also take `version = "IRF"`, and `label_pid5()` labels items with the informant wording. No validity scales for the informant form are on the package's source shelf, and its norms (Markon et al., 2024, Tables A–10 and A–11) are not yet in `pid_norms`, so `validity_pid5()`, `norm_pid5()` and `plot_pid5()` do not take it. Because informant scores carry the full form's column names, `norm_pid5()` and `plot_pid5()` would accept them as `version = "FULL"` without complaint; do not do that, as it compares informant ratings with self-report norms.

## Collecting Responses Online with hitop-form

[hitop-form](https://jmgirard.github.io/hitop-form/) is an online form that shows the PID-5, the PID-5-SF or the PID-5-BF in the browser. It needs no survey platform. The item text, response options and instructions come from this package's JSON export of the form. A study link tells the online form where each participant's answers go: a file saved on the participant's own device, which they send to you, or a web address or Supabase table the link names. A Google Sheet's script is one such web address. You download the sheet or the table as one CSV file. The [Collecting Responses Online](https://jmgirard.github.io/hitop/articles/online-collection.html) article walks the Google Sheet route end to end.

To make a study link, open the [Study Link Builder](https://jmgirard.github.io/hitop-form/link.html). Choose the form, name the study and choose where the responses go. To give a participant identifier, open the closed "Participants and recruiting site" section. Press "Make the link", and the link shows under "Your study link". Send it to the participant.

Whichever route the answers take, the CSV file has five lead columns (`study`, `participant`, `instrument`, `form_build`, `submitted`), a sixth, `item_order`, when the link asked for a random order, two more, `prolific_study` and `prolific_session`, when the link recruits through Prolific, and then one column per item. The item columns are named as this package names them: `pid5_001` to `pid5_220`, `pid5sf_001` to `pid5sf_100`, or `pid5bf_01` to `pid5bf_25`. Each holds the value of the chosen option, 0 to 3. A file the online form saved holds one participant's row; the download of a sheet or a table holds one row per participant.

`read_form_responses()` reads a folder of these files, or a vector of their paths, into one data frame with one row per response row of each file. Read the files of each form in a call of their own. The file below is one the online form saved from the full form. The package installs it as an example, and `system.file()` gives its path.

```{r}
path <- system.file("examples", "responses-pid5.csv", package = "hitop")
responses <- read_form_responses(path)
responses
```

The item columns are integers and follow the form's item order, so pass them to `score_pid5()` by name with the matching `version` (`"FULL"`, `"SF"` or `"BF"`):

```{r}
items <- grep("^pid5_", names(responses), value = TRUE)
score_pid5(responses, items = items, version = "FULL", append = FALSE)
```
