---
title: "Scoring the PID-5"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring the PID-5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| include: false
library(tibble)
```

The Personality Inventory for DSM-5 (PID-5) instrument has 220 items and yields 25 facet scales, 5 domain scales, and 5 validity scales. We can demonstrate the package's functionality using some simulated data.

First, we load the package into memory using the `library()` function. If this doesn't work, make sure you installed the package properly (see the README on [GitHub](https://github.com/jmgirard/hitop)).

```{r}
library(hitop)
```

## Score simulated PID-5 data

The `sim_pid5` dataset is built into the package and can be loaded using the `data()` function. It contains 100 rows (each representing a simulated participant) and 220 columns named `pid5_001` to `pid5_220` (each representing an item from the PID-5).

```{r}
data("sim_pid5")
sim_pid5
```

If your own PID-5 columns are named some other way, `rename_pid5_items()` will rename them to this pattern. By default it reads the item number out of a column already named `pid_1`, `pid_2` and so on -- the spelling this package's own datasets used before they were renamed to match the exports. Setting `method = "text"` instead matches the literal item prompts. Columns it cannot match keep their names and are reported, and a partial rename warns that fewer than 220 items were matched, which suits a study that administered only some of them.

```{r}
old_names <- data.frame(pid_1 = c(0, 1), pid_2 = c(2, 3), age = c(30, 40))

rename_pid5_items(old_names, version = "FULL")
```

To turn these item-level data into scale scores on the 25 facets and 5 domains, we can use the `score_pid5()` function. We will need to tell the function which columns contain our items and which version of the PID this is. There are several ways we can specify the items. First, we can provide the column numbers and use the `:` shortcut. In this tibble, the items are from column 1 to column 220 so we can use `items = 1:220`. I am going to also set `append = FALSE` so that you can quickly see the scale scores. I also can set the version to `"FULL"` (or leave that argument off, as that is the default, shown in the example below) to let it know we are using the full 220-item version.

```{r}
scores <- score_pid5(sim_pid5, items = 1:220, version = "FULL", append = FALSE)
scores
```

If I had instead set `append = TRUE` (or left it off, as that is the default), we would get back the `sim_pid5` tibble with the scale scores added to the end as extra columns. Notice below how we now have 250 columns instead of 220 or 30.

```{r}
scores <- score_pid5(sim_pid5, items = 1:220)
scores
```

Alternatively, we could provide the item column names as a character string. Typing out all 220 item names would be a hassle, but luckily this data named them consistently so we can build the names automatically using `sprintf()`. If we use the "pid5_%03d" format and apply that across the numbers 1 to 220, that will create the column names we need.

```{r}
scores <- score_pid5(sim_pid5, items = sprintf("pid5_%03d", 1:220))
scores
```

There are other useful arguments to the function that you can read about using its documentation by typing the following into your R console: `?score_pid5` or through the [package website](https://jmgirard.github.io/hitop/reference/score_pid5.html).

## Rank Each Person's Highest Facets

Thirty scale columns is more than can be read at a glance. The `rank_scales()` function condenses them: for each row it sorts the columns you select and returns a single string naming the highest-scoring ones, highest first. Here we rank the 25 facet scales — the first 25 columns of the scored tibble, the last 5 being the domains those facets roll up into — and ask for each participant's top 5. Setting `prefix = "pid_"` strips that leading string from the column names, so the string reads as scale names.

```{r rank}
facet_scores <- score_pid5(sim_pid5, items = 1:220, append = FALSE)
rank_scales(facet_scores, scales = 1:25, prefix = "pid_", top = 5, append = FALSE)
```

The `top_scales` column holds one comma-separated string per participant: the names of their five highest-scoring facets, in descending order. Ties are broken by the order of the columns you selected. We passed `append = FALSE` so that the ranked column comes back on its own; leaving that argument at its default returns the scored tibble with the column added to the end instead. Use `name` to call the output column something else and `dir = "low"` to rank from the bottom.

## Scale Reliability

As we compute scale scores, we can also estimate their inter-item reliability using Cronbach's α (alpha) or McDonald's ω (omega total). α is fast and widely used, but it assumes tau-equivalence (all items load equally on a single factor); violations can make α under- or over-estimate reliability. ω is based on a congeneric single-factor model, allowing items to have different loadings and error variances; it typically provides a more accurate reliability estimate for unit-weighted sums. Both assume the scale is essentially unidimensional; α and ω coincide when tau-equivalence holds.

We estimate reliability with the `reliability_pid5()` function, which returns a tibble with one row per scale: its printed name (`Scale`), the stem that names its column in the scored output (`camelCase`), the number of items (`nItems`), and the requested coefficients. By default it computes both `alpha` and `omega`; for the latter, we will need the **lavaan** package installed (set `omega = FALSE` to skip it). Note that, because this is naively simulated data, we would expect the reliability in this example to be poor.

```{r reliability}
#| warning: false
reliability_pid5(
  data = sim_pid5,
  items = sprintf("pid5_%03d", 1:220),
  version = "FULL"
)
```

## Validity Scales for the PID-5

There are also several validity scales that have been developed for the full PID-5, including measures of overreporting, inconsistent responding, and positive impression management. We can use the simulated data to demonstrate the ability of the `validity_pid5()` function to calculate these scores and flag issues. The function arguments will be consistent with what we just learned. Note that, because the data is fake, we would expect there to be lots of validity issues.

```{r x2b}
validity_pid5(sim_pid5, items = 1:220, append = FALSE)
```

## Normative Scores

The `pid_norms` dataset carries the normative score distributions published by Markon et al. (2024). The `norm_pid5()` function looks scored columns up in those tables and returns, for each one, the T score and percentile printed against the nearest tabled raw score. It converts scores rather than computing them, so we hand it the output of `score_pid5()` — and of `validity_pid5()`, if we want those scales converted too.

```{r x2c}
#| message: false
#| warning: false
scored <- score_pid5(sim_pid5, items = 1:220)
scored <- validity_pid5(scored, items = 1:220)
```

For the full form the published tables cover the five domain scales, all 25 facet scales, and three of the validity scales — INC, ORS, and PRD. They do not cover SD-TD at all. Each converted scale gains a `_ptl` column, and those whose tables print T scores also gain a `_t` column. The validity scales are distributed as percentiles only, so they get no `_t` column.

```{r x2d}
norm_pid5(
  scored,
  scores = paste0(
    "pid_",
    c("negativeAffectivity", "detachment", "antagonism", "disinhibition",
      "psychoticism", "INC", "ORS", "PRD")
  ),
  version = "FULL",
  append = FALSE
)
```

The facets convert the same way. Anhedonia, for instance, is an item mean over its eight full-form items, and its table runs from a floor of 0.00 up to a printed 3.84 — past the 3.00 an item mean can actually reach, which is true of most of the facet columns and is a property of the published tables rather than of the conversion:

```{r x2d2}
norm_pid5(scored, scores = "pid_anhedonia", version = "FULL", append = FALSE)
```

Every number returned is a cell of a published table: the nearest printed row is selected and nothing is interpolated. A score that falls outside a printed range is capped to the nearest end rather than extrapolated, and a warning reports how many observations that happened to — the `PRD` sum reaches 66 while its table stops at 55, so it is a common one to see. A scale the tables do not cover returns `NA` in both columns with a warning naming it. Every report this function makes is a warning, so a single `suppressWarnings()` call silences it.

If the items were answered on a four-option response scale that starts somewhere other than 0 — 1 to 4, say — pass that range as `srange` and each score is reconciled to the published 0–3 metric before it is looked up, with a warning naming which scales were adjusted and which were left alone. The per-scale formulas are given in `?norm_pid5`.

## Profile Plots

Once a respondent's scores are normed, `plot_pid5()` draws them as a profile against the published metric. It takes one respondent — a profile plot shows one person — so we norm the whole dataset and hand it a single row.

```{r x2e}
#| message: false
#| warning: false
domains <- paste0(
  "pid_",
  c("negativeAffectivity", "detachment", "antagonism", "disinhibition",
    "psychoticism")
)
normed <- norm_pid5(scored, scores = domains, version = "FULL")
```

```{r x2f}
#| fig.width: 7
#| fig.height: 4
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
plot_pid5(normed[1, ], version = "FULL")
```

The dashed line marks T = 50, the normative sample's mean, and the score axis spans the range the published tables actually print for these scales — so the axis does not rescale from respondent to respondent and two profiles are directly comparable. Nothing on the plot says whether a score is high, low, or concerning: {hitop} presents scores against norms and leaves the interpreting to you.

Passing `level = "facet"` plots all 25 facets instead, grouped into a panel per domain. The APA key ties three facets to each domain; the remaining ten define no domain and are grouped separately rather than dropped.

```{r x2g}
#| fig.width: 7.5
#| fig.height: 9
#| message: false
#| warning: false
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
facets <- paste0("pid_", pid_scales[["FULL"]]$camelCase)
normed_facets <- norm_pid5(scored, scores = facets, version = "FULL")
plot_pid5(normed_facets[1, ], version = "FULL", level = "facet")
```

Set `metric = "percentile"` for a percentile axis instead of T scores. `norm_pid5()` returns percentiles as a proportion; the plot multiplies them by 100 so the axis reads 0–100.

```{r x2h}
#| fig.width: 7
#| fig.height: 4
#| eval: !expr rlang::is_installed("ggplot2", version = "3.4.0")
plot_pid5(normed[1, ], version = "FULL", metric = "percentile")
```

The result is an ordinary ggplot object, so you can restyle it with any ggplot2 layer — `+ ggplot2::labs(title = ...)`, a different theme, and so on.
