---
title: "HiTOP-HSUM Instrument"
output: rmarkdown::html_vignette
toc: false
vignette: >
  %\VignetteIndexEntry{HiTOP-HSUM Instrument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the HiTOP Harmful Substance Use Module (HiTOP-HSUM) instrument resources page. This questionnaire contains a total of 650 possible items across 3 assessment stages. Complex skip/display logic is used, such that the vast majority of respondents will only see a subset of possible items. Here you can download a DOCX summary file that can be printed and shared with IRB as well as ready-to-use imports for both Qualtrics and REDCap (both of which already implement the complex skip/display logic mentioned above). Scoring functions are still under development but should be available soon.

```{r setup, include=FALSE}
source("_download-helpers.R")
```

```{=html}
<h3 class="mt-4 mb-3">Ready-to-Use Downloads</h3>
<p class="text-muted">Choose the format that best fits your immediate research needs.</p>
```

```{r downloads, echo=FALSE, results='asis'}
download_cards("HiTOP-HSUM", list(
  dl_card(
    "📄", "Printable Document",
    "Use these Microsoft Word documents for an overview or sending to the IRB.",
    dl_link("English (A4 Paper)", "../downloads/hitophsum_A4.docx"),
    dl_link("English (US Paper)", "../downloads/hitophsum_US.docx")
  ),
  dl_card(
    "📊", "Qualtrics Import",
    "Use this survey file to easily import the instrument directly into your Qualtrics surveys.",
    dl_link("English (QSF File)", "../downloads/hitophsum_qualtrics.qsf"),
    ref_link("Import Instructions", "../articles/import-instructions.html#qualtrics-qsf")
  ),
  dl_card(
    "🏥", "REDCap Import",
    "Use this compressed archive file to import the instrument as a new instrument in your REDCap project.",
    dl_link("English (ZIP File)", "../downloads/hitophsum_redcap.zip"),
    ref_link("Import Instructions", "../articles/import-instructions.html#redcap-zip")
  )
))
```

Scoring functions and example data coming soon.

```{=html}
<hr class="mt-5 mb-4">

<h3 class="mb-3">Explore the R Package Features</h3>
<p class="text-muted">Scoring for the HiTOP-HSUM is still under development, but the <code>hitop</code> package can already build the distributed files above from scratch, so you can adapt them before fielding the instrument.</p>

<div class="row mt-4">
  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">🛠️ Custom File Generation</h5>
        <p class="card-text">Need to change the formatting, adapt the instructions, or translate the text? Build customized DOCX and REDCap files programmatically. The Qualtrics survey file is distributed as a prebuilt download only.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/generate_docx_hitophsum.html" class="btn btn-sm btn-outline-secondary">Printable</a>
          <a href="../reference/generate_redcap_hitophsum.html" class="btn btn-sm btn-outline-secondary">REDCap</a>
        </div>
      </div>
    </div>
  </div>
</div>
```

```{r versions, echo=FALSE, results='asis'}
versions_section("HiTOP-HSUM")
```
