---
title: "PID-5 Child Form Instrument"
output: rmarkdown::html_vignette
toc: false
vignette: >
  %\VignetteIndexEntry{PID-5 Child Form Instrument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the resources page for the Personality Inventory for DSM-5 child form for ages 11 to 17 (Krueger et al., 2013). The child completes this 220-item questionnaire about themselves. Its items, their order and their scoring key are those of the adult PID-5. Its instructions put "right" and "wrong" in quotation marks, and the Word forms here carry the APA notice that the child form prints. The APA form also labels its instructions "Instructions to the child receiving care", which these forms leave out, as the adult forms leave out their label. Here you can download ready-to-use versions of the instrument or use the `hitop` R package to customize, score, and analyze your data.

```{r setup, include=FALSE}
source("_download-helpers.R")
```

```{=html}
<h3 class="mt-4 mb-3">Ready-to-Use Downloads</h3>
<p class="text-muted">Choose the format that best fits your immediate research needs.</p>
```

```{r downloads, echo=FALSE, results='asis'}
download_cards("PID-5 Child", list(
  dl_card(
    "📄", "Printable Document",
    "Use these Microsoft Word documents for printing, paper administration, or sending to the IRB.",
    dl_link("English (A4 Paper)", "../downloads/pid5child_A4.docx"),
    dl_link("English (US Paper)", "../downloads/pid5child_US.docx")
  ),
  dl_card(
    "📊", "Qualtrics Import",
    "Use this specially formatted text file to easily import the instrument directly into your Qualtrics surveys.",
    dl_link("English (TXT File)", "../downloads/pid5child_qualtrics.txt"),
    ref_link("Import Instructions", "../articles/import-instructions.html#qualtrics-txt")
  ),
  dl_card(
    "🏥", "REDCap Import",
    "Use this compressed archive file to import the instrument as a new instrument in your REDCap project.",
    dl_link("English (ZIP File)", "../downloads/pid5child_redcap.zip"),
    ref_link("Import Instructions", "../articles/import-instructions.html#redcap-zip")
  )
))
```

```{=html}
<hr class="mt-5 mb-4">

<h3 class="mb-3">Explore the R Package Features</h3>
<p class="text-muted">The <code>hitop</code> package provides a toolkit for working with the PID-5 child form. If you are comfortable using R, you can use these functions in your workflow.</p>

<div class="row mt-4">
  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">📋 Instrument Information</h5>
        <p class="card-text">Access the item dictionary and the facet keys directly from the package namespace. The child form uses the adult full form's items and keys.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/pid_items.html" class="btn btn-sm btn-outline-secondary">Items</a>
          <a href="../reference/pid_scales.html" class="btn btn-sm btn-outline-secondary">Facets</a>
          <a href="../reference/pid_domains.html" class="btn btn-sm btn-outline-secondary">Domains</a>
        </div>
      </div>
    </div>
  </div>

  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">🛠️ Custom File Generation</h5>
        <p class="card-text">Need to change the paper size, font, title or page breaks? Build customized DOCX, Qualtrics, and REDCap files programmatically.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/generate_docx_pid5child.html" class="btn btn-sm btn-outline-secondary">Printable</a>
          <a href="../reference/generate_qualtrics_pid5child.html" class="btn btn-sm btn-outline-secondary">Qualtrics</a>
          <a href="../reference/generate_redcap_pid5child.html" class="btn btn-sm btn-outline-secondary">REDCap</a>
        </div>
      </div>
    </div>
  </div>

  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">🧮 Scoring & Reliability</h5>
        <p class="card-text">Score the facets and domains with <code>version = "FULL"</code>, as for the adult form, or compute alpha and omega for each facet. The package has no child norms, so do not norm child scores against the adult tables.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/score_pid5.html" class="btn btn-sm btn-outline-secondary">Score PID-5</a>
          <a href="../reference/reliability_pid5.html" class="btn btn-sm btn-outline-secondary">Reliability</a>
        </div>
      </div>
    </div>
  </div>
</div>
```

```{r versions, echo=FALSE, results='asis'}
versions_section("PID-5 Child")
```
