---
title: "HiTOP-HSUM Instrument"
output: rmarkdown::html_vignette
toc: false
vignette: >
  %\VignetteIndexEntry{HiTOP-HSUM Instrument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the HiTOP Harmful Substance Use Module (HiTOP-HSUM) instrument resources page. This questionnaire contains a total of 650 possible items across 3 assessment stages. Complex skip/display logic is used, such that the vast majority of respondents will only see a subset of possible items. Here you can download a DOCX summary file that can be printed and shared with IRB as well as ready-to-use imports for both Qualtrics and REDCap (both of which already implement the complex skip/display logic mentioned above). Scoring functions are still under development but should be available soon.

```{r setup, include=FALSE}
source("_download-helpers.R")
```

```{=html}
<h3 class="mt-4 mb-3">Ready-to-Use Downloads</h3>
<p class="text-muted">Choose the format that best fits your immediate research needs.</p>
```

```{r downloads, echo=FALSE, results='asis'}
download_cards("HiTOP-HSUM", list(
  dl_card(
    "📄", "Printable Document",
    "Use these Microsoft Word documents for an overview or sending to the IRB.",
    dl_link("English (A4 Paper)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitophsum_A4.docx"),
    dl_link("English (US Paper)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitophsum_US.docx")
  ),
  dl_card(
    "📊", "Qualtrics Import",
    "Use this survey file to easily import the instrument directly into your Qualtrics surveys.",
    dl_link("English (QSF File)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitophsum_qualtrics.qsf")
  ),
  dl_card(
    "🏥", "REDCap Import",
    "Use this compressed archive file to import the instrument as a new instrument in your REDCap project.",
    dl_link("English (ZIP File)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitophsum_redcap.zip"),
    ref_link("Import Instructions", "../reference/generate_redcap_hitophsum.html#details")
  )
))
```

Scoring functions and example data coming soon.

```{r versions, echo=FALSE, results='asis'}
versions_section("HiTOP-HSUM")
```
