---
title: "HiTOP-SR Instrument"
output: rmarkdown::html_vignette
toc: false
vignette: >
  %\VignetteIndexEntry{HiTOP-SR Instrument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to the HiTOP Self-Report (HiTOP-SR) instrument resources page. This questionnaire contains 405 items and can be used to score 76 scales and 17 subscales indexing different aspects of psychopathology. Here you can download ready-to-use versions of the instrument or explore the powerful tools provided by the `hitop` R package to customize, score, and analyze your data.

```{r setup, include=FALSE}
source("_download-helpers.R")
```

```{=html}
<h3 class="mt-4 mb-3">Ready-to-Use Downloads</h3>
<p class="text-muted">Choose the format that best fits your immediate research needs.</p>
```

```{r downloads, echo=FALSE, results='asis'}
download_cards("HiTOP-SR", list(
  dl_card(
    "📄", "Printable Document",
    "Use these Microsoft Word documents for printing, paper administration, or sending to the IRB.",
    dl_link("English (A4 Paper)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitopsr_A4.docx"),
    dl_link("English (US Paper)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitopsr_US.docx")
  ),
  dl_card(
    "📊", "Qualtrics Import",
    "Use this specially formatted text file to easily import the instrument directly into your Qualtrics surveys.",
    dl_link("English (TXT File)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitopsr_qualtrics.txt")
  ),
  dl_card(
    "🏥", "REDCap Import",
    "Use this compressed archive file to import the instrument as a new instrument in your REDCap project.",
    dl_link("English (ZIP File)", "https://github.com/jmgirard/hitop/raw/main/inst/extdata/hitopsr_redcap.zip"),
    ref_link("Import Instructions", "../reference/generate_redcap_hitopsr.html#details")
  )
))
```

```{=html}
<hr class="mt-5 mb-4">

<h3 class="mb-3">Explore the R Package Features</h3>
<p class="text-muted">The <code>hitop</code> package provides a complete toolkit for working with the HiTOP-SR. If you are comfortable using R, you can leverage these functions to streamline your workflow.</p>

<div class="row mt-4">
  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">📋 Instrument Information</h5>
        <p class="card-text">Access the complete item dictionary, structural hierarchy, and subscale definitions directly from the package namespace.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/hitopsr_items.html" class="btn btn-sm btn-outline-secondary">Items</a>
          <a href="../reference/hitopsr_scales.html" class="btn btn-sm btn-outline-secondary">Scales</a>
          <a href="../reference/hitopsr_subscales.html" class="btn btn-sm btn-outline-secondary">Subscales</a>
          <a href="../articles/scales-hitopsr.html" class="btn btn-sm btn-outline-secondary">Definitions</a>
        </div>
      </div>
    </div>
  </div>

  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">🛠️ Custom File Generation</h5>
        <p class="card-text">Need to change the formatting, adapt the instructions, or translate the text? Build customized DOCX, Qualtrics, and REDCap files programmatically.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/generate_docx_hitopsr.html" class="btn btn-sm btn-outline-secondary">Printable</a>
          <a href="../reference/generate_qualtrics_hitopsr.html" class="btn btn-sm btn-outline-secondary">Qualtrics</a>
          <a href="../reference/generate_redcap_hitopsr.html" class="btn btn-sm btn-outline-secondary">REDCap</a>
        </div>
      </div>
    </div>
  </div>

  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">🧮 Scoring & Reliability</h5>
        <p class="card-text">Calculate subscales, handle reverse-coding, and generate reliability metrics in one step, or access the underlying internal consistency functions directly.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/score_hitopsr.html" class="btn btn-sm btn-outline-secondary">Score HiTOP-SR</a>
          <a href="../reference/calc_alpha.html" class="btn btn-sm btn-outline-secondary">Alpha Reliability</a>
          <a href="../reference/calc_omega.html" class="btn btn-sm btn-outline-secondary">Omega Reliability</a>
        </div>
      </div>
    </div>
  </div>

  <div class="col-md-6 mb-4">
    <div class="card h-100 shadow-sm">
      <div class="card-body">
        <h5 class="card-title">📊 Example Datasets</h5>
        <p class="card-text">Test your analysis pipelines using built-in datasets, including a real validation sample and simulated respondent data.</p>
        <div class="d-flex flex-wrap gap-2">
          <a href="../reference/ku_hitopsr.html" class="btn btn-sm btn-outline-secondary">Real Data</a>
          <a href="../reference/sim_hitopsr.html" class="btn btn-sm btn-outline-secondary">Simulated Data</a>
        </div>
      </div>
    </div>
  </div>
</div>
```

```{r versions, echo=FALSE, results='asis'}
versions_section("HiTOP-SR")
```
