---
title: "Building HiTOP-SR Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building HiTOP-SR Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tibble)
```

The full HiTOP-SR asks 405 items and returns 76 scale scores, which is more
than many studies need and more than many participants will sit through. A
**module** is a chosen set of the instrument's scales, administered and scored
on its own.

This article walks the whole module workflow end to end: choosing which scales
to include, describing the module, generating the files you will actually field,
and scoring the data that comes back. For scoring a *complete* administration,
see [Scoring the HiTOP-SR](hitopsr_scoring.html) instead.

```{r}
library(hitop)
```

## Choosing Scales

Start from the menu. `available_scales()` lists every scale the instrument
offers, along with the camelCase stem that will name its scored output column
and the number of items it contributes.

```{r}
scale_menu <- available_scales("hitopsr")
scale_menu
```

Both name columns are accepted when you build a module, so you can write scale
names the way they are printed on the form (`"Antisocial Behavior"`) or the way
they appear in scored output (`"antisocialBehavior"`).

A module is built from these 76 scales only. The instrument also defines 17
subscales (see `hitopsr_subscales`), but a subscale is not a unit you can select
here — `hitop_module()` rejects a subscale name as unknown. Ask for the scale
that contains it instead.

Because this is an ordinary tibble, you can also use it to plan the length of
your module. For example, to see the shortest scales:

```{r}
scale_menu[order(scale_menu$nItems), ][1:5, ]
```

## Building the Module

`hitop_module()` turns a set of scale names into a validated description of the
module. It resolves the names, works out which items those scales need, and
records the reverse-keying flags that scoring will require.

```{r}
four_scale <- hitop_module(
  instrument = "hitopsr",
  scales = c(
    "Agoraphobia", "Appetite Loss",
    "Antisocial Behavior", "Romantic Disinterest"
  )
)
four_scale
```

The printed summary reports how many items the module needs. The description
itself is a plain list, so you can look inside it:

```{r}
four_scale$camelCase
four_scale$nItems
four_scale$items
```

Note that `$items` holds the **original** HiTOP-SR item numbers. That is true
of the descriptor whatever a generator later prints: item 42 of the full
instrument is item 42 here, which is what lets data collected with a module be
scored against the full instrument's key.

## Generating the Instrument

Each of the three HiTOP-SR generators takes a `module` argument. Pass the
description and you get a form containing only that module's items.

The three do not number those items alike, on purpose. The Word form is what a
participant fills in on paper, so `generate_docx_hitopsr()` numbers it `1` to
`n` down the page -- a four-scale module reads 1, 2, 3 rather than 7, 42, 213.
Pass `renumber = FALSE` to keep the original numbers instead. In the Qualtrics
and REDCap exports an item number names a collected data column, so those keep
the original HiTOP-SR numbers always; renumbering them would rename variables
in dictionaries already in the field.

`generate_docx_hitopsr()` also takes `randomize = TRUE`, which prints the items
in a random order. The page is still numbered `1` to `n`, and the scoring page
gains a crosswalk from each printed number back to its original HiTOP-SR
number, so a shuffled form can be scored from the paper alone. Call
`set.seed()` beforehand to make an order reproducible, and read the order back
from the `item_order` attribute of the returned path.

Here we write all three formats into a temporary folder; in your own work you
would give a real path or let the default filename land in your working
directory.

```{r}
#| message: false
outdir <- tempdir()

docx_file <- generate_docx_hitopsr(
  file = file.path(outdir, "hitopsr_module.docx"),
  module = four_scale
)

qualtrics_file <- generate_qualtrics_hitopsr(
  file = file.path(outdir, "hitopsr_module.txt"),
  module = four_scale
)

redcap_file <- generate_redcap_hitopsr(
  file = file.path(outdir, "hitopsr_module.zip"),
  module = four_scale
)
```

```{r}
basename(c(docx_file, qualtrics_file, redcap_file))
```

The DOCX file is a paper form ready to print, the `.txt` file imports into
Qualtrics as an advanced-format question block, and the `.zip` file imports into
REDCap as an instrument. See [Importing Instruments into Qualtrics and
REDCap](import-instructions.html) for the import steps themselves.

## Selecting the Collected Columns

Once the data comes back, it has only the module's item columns — plus whatever
else your platform recorded, such as participant identifiers and demographics.
Both scoring functions need you to name those item columns.

We will stand in for module-collected data using `ku_hitopsr`, the package's
real example dataset. Its first two columns are `participant` and `biosex`, and
its 405 item columns are named `hsr001` through `hsr405`:

```{r}
data("ku_hitopsr")
names(ku_hitopsr)[1:5]
```

**Select the item columns by name, never by position.** The numbers in
`four_scale$items` are item numbers, not column numbers. They coincide only in
a data frame that is exactly the 405 items in order — and this one is not,
because two other columns come first. Build the column names instead:

```{r}
item_cols <- sprintf("hsr%03d", four_scale$items)
collected <- ku_hitopsr[item_cols]
ncol(collected)
```

That width matches the module's own item count, which is the check worth
running before you score anything:

```{r}
ncol(collected) == four_scale$nItems
```

Indexing by `four_scale$items` directly would have selected columns 42, 66, 68,
and so on *by position*, which in this frame are the wrong items — shifted by
the two leading columns:

```{r}
names(ku_hitopsr[four_scale$items])[1:4]
```

Those are items 40, 64, 66, and 107 masquerading as the module's items. Nothing
would have failed loudly; the scores would simply have been wrong. If your item
columns carry other names, `rename_hitopsr_items()` can rename them to standard
item numbers first; its `prefix` argument sets the stem those names are built
from.

## Scoring and Reliability

Hand the same description back through the `module` argument. Without it,
`score_hitopsr()` expects a full administration and stops on the item count
("Expected 405 items but got 21"); with it, the `items` argument names the
columns you actually have, in instrument order, and `module` tells the function
which scales they belong to.

```{r}
module_scores <- score_hitopsr(
  data = collected,
  items = names(collected),
  module = four_scale,
  append = FALSE
)
module_scores
```

Only the module's scales come back, in the order they appear in
`hitopsr_scales`. The values are exactly what a full administration would have
produced for those scales — a scale score depends only on its own items, so
dropping the other 72 scales' columns cannot move it:

```{r}
full_scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr%03d", 1:405),
  append = FALSE
)
all.equal(module_scores, full_scores[names(module_scores)])
```

`reliability_hitopsr()` takes the same argument and returns one row per module
scale.

```{r}
reliability_hitopsr(
  data = collected,
  items = names(collected),
  module = four_scale,
  omega = FALSE
)
```

Reliability is worth checking on your own module rather than assumed. In this
sample the four coefficients run from modest down to negative: Romantic
Disinterest returns a negative α, which is not a small number but a sign that
its items covary near zero or oppositely in these data, so a mean of them is
not measuring one thing. Short scales make α less stable, but shortness alone
does not produce a negative value. Set `omega = FALSE` to skip McDonald's ω,
which needs the **lavaan** package.

The `srange`, `prefix`, `missing`, `calc_se`, and `append` arguments all behave
exactly as they do for a full administration; see [Scoring the
HiTOP-SR](hitopsr_scoring.html) for those.
