---
title: "Building HiTOP-SR Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building HiTOP-SR Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tibble)
```

The full HiTOP-SR asks 405 items and returns 76 scale scores, which is more
than many studies need and more than many participants will sit through. A
**module** is a chosen set of the instrument's scales, administered and scored
on its own.

This article walks the whole module workflow end to end: choosing which scales
to include, describing the module, generating the files you will actually field,
and scoring the data that comes back. For scoring a *complete* administration,
see [Scoring the HiTOP-SR](hitopsr_scoring.html) instead.

```{r}
library(hitop)
```

## Choosing Scales

Start from the menu. `available_scales()` lists every scale the instrument
offers, along with the camelCase stem that will name its scored output column,
the number of items it contributes, and a brief clinician-facing definition
(`Brief`, printed here truncated; `hitopsr_definitions` carries it in full).

```{r}
scale_menu <- available_scales("hitopsr")
scale_menu
```

Both name columns are accepted when you build a module, so you can write scale
names the way they are printed on the form (`"Antisocial Behavior"`) or the way
they appear in scored output (`"antisocialBehavior"`).

A module is built from these 76 scales only. The instrument also defines 17
subscales (see `hitopsr_subscales`), but a subscale is not a unit you can select
here — `hitop_module()` rejects a subscale name as unknown. Ask for the scale
that contains it instead.

Because this is an ordinary tibble, you can also use it to plan the length of
your module. For example, to see the shortest scales:

```{r}
scale_menu[order(scale_menu$nItems), ][1:5, ]
```

## Building the Module

`hitop_module()` turns a set of scale names into a validated description of the
module. It resolves the names, works out which items those scales need, and
records the reverse-keying flags that scoring will require.

```{r}
four_scale <- hitop_module(
  instrument = "hitopsr",
  scales = c(
    "Agoraphobia", "Appetite Loss",
    "Antisocial Behavior", "Romantic Disinterest"
  )
)
four_scale
```

The printed summary reports how many items the module needs. The description
itself is a plain list, so you can look inside it:

```{r}
four_scale$camelCase
four_scale$nItems
four_scale$items
```

Note that `$items` holds the **original** HiTOP-SR item numbers. That is true
of the descriptor whatever a generator later prints: item 42 of the full
instrument is item 42 here, which is what lets data collected with a module be
scored against the full instrument's key.

## Generating the Instrument

Each of the three HiTOP-SR generators takes a `module` argument. Pass the
description and you get a form containing only that module's items.

The three do not number those items alike, on purpose. The Word form is what a
participant fills in on paper, so `generate_docx_hitopsr()` numbers it `1` to
`n` down the page -- a four-scale module reads 1, 2, 3 rather than 7, 42, 213.
Pass `renumber = FALSE` to keep the original numbers instead. In the Qualtrics
and REDCap exports an item number names a collected data column, so those keep
the original HiTOP-SR numbers always; renumbering them would rename variables
in dictionaries already in the field.

A Word form also says on its face that it is a module. `generate_docx_hitopsr()`
heads a module form `HiTOP-SR Module (v1.0)` where it heads a full-instrument
form `HiTOP-SR (v1.0)`. Pass `title =` to print something else — a study name,
say — and that string is used verbatim on either kind of form.

One combination the Word generator refuses is `include_subscales = TRUE`
together with `module`: a subscale can draw items from scales outside the
module, so its scoring row would list items the form does not contain. The call
aborts rather than print a row that cannot be scored.

`generate_docx_hitopsr()` also takes `randomize = TRUE`, which prints the items
in a random order. A renumbered module form — the default — is still numbered
`1` to `n` and carries a crosswalk from each printed number back to its original
HiTOP-SR number, printed whether or not `include_scoring` appends the key, so it
can be scored from the paper alone. That is the only case the crosswalk is
printed in. Under `renumber = FALSE` there is none, because the printed numbers
already are the original ones; and shuffling the full instrument prints none
either, since 405 pairs would fill a page. In both of those cases read the order
from the `item_order` attribute of the returned path, or save it with
`descriptor =` below. Call `set.seed()` beforehand to make an order
reproducible.

One thing to watch: by default `score_hitopsr()` expects a module's items in
ascending **original** order, not the order a shuffled form prints them in.
Data entered straight off a shuffled form has its columns in the order the
form printed: the first column holds the answer to the paper's item 1, the
second the answer to item 2, and so on. Score such columns with
`layout = "printed"` and a module that carries the printed order on its
`item_order` attribute, which is what the descriptor saved with `descriptor =`
below gives you. The function then puts the columns back into instrument
order itself. Scoring printed-order columns under the default layout returns
wrong scale scores with no error raised, and scoring columns already in
instrument order under `layout = "printed"` scrambles what was right.

Here we write all three formats into a temporary folder; in your own work you
would give a real path or let the default filename land in your working
directory.

```{r}
#| message: false
outdir <- tempdir()

docx_file <- generate_docx_hitopsr(
  file = file.path(outdir, "hitopsr_module.docx"),
  module = four_scale
)

qualtrics_file <- generate_qualtrics_hitopsr(
  file = file.path(outdir, "hitopsr_module.txt"),
  module = four_scale
)

redcap_file <- generate_redcap_hitopsr(
  file = file.path(outdir, "hitopsr_module.zip"),
  module = four_scale
)
```

```{r}
basename(c(docx_file, qualtrics_file, redcap_file))
```

The DOCX file is a paper form ready to print, the `.txt` file imports into
Qualtrics as an advanced-format question block, and the `.zip` file imports into
REDCap as an instrument. See [Importing Instruments into Qualtrics and
REDCap](import-instructions.html) for the import steps themselves.

## Selecting the Collected Columns

Once the data comes back, it has only the module's item columns — plus whatever
else your platform recorded, such as participant identifiers and demographics.
Both scoring functions need you to name those item columns.

We will stand in for module-collected data using `ku_hitopsr`, the package's
real example dataset. Its first two columns are `participant` and `biosex`, and
its 405 item columns are named `hsr_001` through `hsr_405`:

```{r}
data("ku_hitopsr")
names(ku_hitopsr)[1:5]
```

**Select the item columns by name, never by position.** The numbers in
`four_scale$items` are item numbers, not column numbers. They coincide only in
a data frame that is exactly the 405 items in order — and this one is not,
because two other columns come first. Build the column names instead:

```{r}
item_cols <- sprintf("hsr_%03d", four_scale$items)
collected <- ku_hitopsr[item_cols]
ncol(collected)
```

That width matches the module's own item count, which is the check worth
running before you score anything:

```{r}
ncol(collected) == four_scale$nItems
```

Indexing by `four_scale$items` directly would have selected columns 42, 66, 68,
and so on *by position*, which in this frame are the wrong items — shifted by
the two leading columns:

```{r}
names(ku_hitopsr[four_scale$items])[1:4]
```

Those are items 40, 64, 66, and 107 masquerading as the module's items. Nothing
would have failed loudly; the scores would simply have been wrong. If your item
columns carry other names, `rename_hitopsr_items()` can rename them to standard
item numbers first; its `prefix` argument sets the stem those names are built
from.

## Scoring and Reliability

Hand the same description back through the `module` argument. Without it,
`score_hitopsr()` expects a full administration and stops on the item count
("Expected 405 items but got 21"); with it, the `items` argument names the
columns you actually have, in instrument order, and `module` tells the function
which scales they belong to.

```{r}
module_scores <- score_hitopsr(
  data = collected,
  items = names(collected),
  module = four_scale,
  append = FALSE
)
module_scores
```

Only the module's scales come back, in the order they appear in
`hitopsr_scales`. The values are exactly what a full administration would have
produced for those scales — a scale score depends only on its own items, so
dropping the other 72 scales' columns cannot move it:

```{r}
full_scores <- score_hitopsr(
  data = ku_hitopsr,
  items = sprintf("hsr_%03d", 1:405),
  append = FALSE
)
all.equal(module_scores, full_scores[names(module_scores)])
```

`reliability_hitopsr()` takes the same argument and returns one row per module
scale.

```{r}
reliability_hitopsr(
  data = collected,
  items = names(collected),
  module = four_scale,
  omega = FALSE
)
```

Reliability is worth checking on your own module rather than assumed. In this
sample the four coefficients run from modest down to negative: Romantic
Disinterest returns a negative α, which is not a small number but a sign that
its items covary near zero or oppositely in these data, so a mean of them is
not measuring one thing. Short scales make α less stable, but shortness alone
does not produce a negative value. Set `omega = FALSE` to skip McDonald's ω,
which needs the **lavaan** package.

The `srange`, `prefix`, `missing`, `calc_se`, and `append` arguments all behave
exactly as they do for a full administration; see [Scoring the
HiTOP-SR](hitopsr_scoring.html) for those. `calc_se` is **deprecated** there and
here alike — it, and the `_se` columns it adds, will be removed in a future
release, and `interval_hitopsr()` replaces it.

## Saving the Module Beside the Form

Everything above depends on still having `four_scale` when the data comes back.
Months later, in a fresh session, rebuilding it means retyping every scale name
— and a typo scores the wrong scales while a forgotten scale scores none.

`write_module()` saves the description to a small JSON file you can keep beside
the forms you generated:

```{r}
descriptor <- write_module(four_scale, file.path(outdir, "hitopsr_module.json"))
cat(readLines(descriptor), sep = "\n")
```

You do not have to remember to call it. Each of the three generators takes a
`descriptor` argument, so one call writes both the form and the file that
scores it:

```{r}
#| message: false
generate_docx_hitopsr(
  file = file.path(outdir, "hitopsr_module2.docx"),
  module = four_scale,
  descriptor = file.path(outdir, "hitopsr_module2.json")
)
```

A call that passes no `module` writes a descriptor naming every scale, so a
full administration is described too. And on a shuffled Word form the
descriptor also records the printed order, which is the one thing a
whole-instrument form gives you nowhere else — no crosswalk is printed for one:

```{r}
#| message: false
set.seed(42)
shuffled_descriptor <- file.path(outdir, "hitopsr_shuffled.json")
generate_docx_hitopsr(
  file = file.path(outdir, "hitopsr_shuffled.docx"),
  module = four_scale,
  randomize = TRUE,
  descriptor = shuffled_descriptor
)

shuffled <- read_module(shuffled_descriptor)
attr(shuffled, "item_order")
```

Those are the original HiTOP-SR item numbers in the order the page printed
them. Responses entered off that form have their columns in the order the form
printed, the first column holding the paper's item 1. Score them with
`layout = "printed"` and the module that carries the order:

```{r}
#| eval: false
score_hitopsr(collected, items = seq_along(collected), module = shuffled,
              layout = "printed")
```

The scoring function puts the columns back into instrument order itself.
Columns already in instrument order take the default, `layout = "instrument"`.

The file is plain text, so you can read it, edit it, and send it to a
collaborator. `read_module()` turns it back into a module:

```{r}
reloaded <- read_module(descriptor)
reloaded
```

What comes back is not simply what the file said. The file records scale
*names*; the items and their reverse-keying flags are rebuilt from this
package's own tables, so a descriptor can never introduce a scoring key of its
own. The `items` the file records are checked against that rebuild, and a
disagreement stops with an error rather than scoring quietly — which is what
you want if the file was written by an older version of the package whose
tables have since changed.

Hand the reloaded module to `score_hitopsr()` exactly as you would the original:

```{r}
identical(
  score_hitopsr(collected, items = names(collected), module = reloaded, append = FALSE),
  module_scores
)
```

### Scoring a REDCap or Qualtrics export without naming the columns

A descriptor written by `generate_redcap_hitopsr()` or
`generate_qualtrics_hitopsr()` also records the names that the generated file
assigns to the items. For REDCap these are the data dictionary's field names.
For Qualtrics they are the questions' `[[ID:]]` values, such as `HSR_066`. A
data export with the item variable names as column headers uses these names.
`read_module()` returns them on the module's `columns` attribute. The REDCap
case looks like this:

```{r}
#| message: false
redcap_descriptor <- file.path(outdir, "hitopsr_redcap.json")
generate_redcap_hitopsr(
  file = file.path(outdir, "hitopsr_module.zip"),
  module = four_scale,
  descriptor = redcap_descriptor
)

from_redcap <- read_module(redcap_descriptor)
head(attr(from_redcap, "columns"))
```

With that module, leave out `items`. The scoring functions take the names from
the module:

```{r}
identical(
  score_hitopsr(collected, module = from_redcap, append = FALSE),
  module_scores
)
```

`reliability_hitopsr()` does the same. Export Qualtrics data as numeric values,
not choice text, so there are numbers to score. Pass `items` for data whose columns carry
other names, such as a Qualtrics export made with "Use internal IDs in header",
or a question renamed after import. A supplied `items` is always used. A module with no `columns`, such as one read
from a Word form's descriptor, still needs `items`.

## Collecting Responses Online with hitop-form

The Word, Qualtrics and REDCap files above each need a platform of their own.
[hitop-form](https://jmgirard.github.io/hitop-form/) is a third route that
needs none: a web page that shows a HiTOP-SR module (or the full HiTOP-SR or
HiTOP-BR) in the browser. A study link tells it where each participant's
answers go: a file saved on the participant's own device, which they send to
you, or a store you name in the link, such as a Google Sheet or a Supabase
table, which you download as one CSV file. The
[Collecting Responses Online](online-collection.html) article walks the
Google Sheet route end to end, from deploying the sheet's script to scoring.

The page reads the same descriptor `write_module()` writes. Open the page's
[link builder](https://jmgirard.github.io/hitop-form/link.html), paste the
descriptor in, name the study, choose where the responses go, and it gives
you one link to send to participants. A descriptor from a shuffled Word form
carries that form's order, and the page shows the items in it. The builder's
"Show the items in a random order" box asks for a new order per participant
instead; the descriptor's order is then not followed.

Whichever route the answers take, the CSV file has five lead columns
(`study`, `participant`, `instrument`, `form_build`, `submitted`) and then one
column per item, named by the instrument's file stem and the item number and
in the order the page showed them. Under the random order, a sixth lead
column, `item_order`, records the order shown, and the item columns are in
the order the descriptor's `items` field lists them, which `write_module()`
writes ascending. A file the page
saved holds one participant's row; a store's download holds one row per
participant.

`read_form_responses()` reads those files. Give it the folder the files are
in, or a vector of their paths. The file below is one the page saved from a
two-scale module with a shuffled order. The package installs it as an example,
and `system.file()` gives its path:

```{r}
path <- system.file("examples", "responses-module-shuffled.csv", package = "hitop")
responses <- read_form_responses(path)
responses
```

Each row of each file is a row of the result, the lead columns typed
(`form_build` is a date, `submitted` a UTC date-time), and every item column
an integer. A folder of many files reads the same way, file by file in path
order. Every file must hold the same item columns in the same order; a
folder mixing two forms stops with an error that names the files that
differ, so read each form's files in a call of their own.

Score the item columns through the module the page showed. Without the
random order, the columns are in
the order the page showed the items, so this is `layout = "printed"` with the
descriptor that recorded that order, as in the shuffled Word form above. (A
file saved under the random order has its columns in instrument order, so it
takes the default layout, as the first scoring call in this article does.) Name
the item columns by position here rather than by name: under
`layout = "printed"` the names carry item numbers out of instrument order, and
naming them would trip the check that warns of a misordered mapping.

```{r}
form_module <- read_module(
  system.file("examples", "module-shuffled.json", package = "hitop")
)
item_cols <- grep("^hitopsr_", names(responses), value = TRUE)
score_hitopsr(
  responses,
  items = match(item_cols, names(responses)),
  module = form_module,
  layout = "printed",
  append = FALSE
)
```

A full HiTOP-SR or HiTOP-BR saved by the page reads the same way and scores
with `score_hitopsr()` or `score_hitopbr()` alone, with `items` naming the
item columns and no module at all.
