---
title: "Collecting Responses Online with hitop-form"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Collecting Responses Online with hitop-form}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[hitop-form](https://jmgirard.github.io/hitop-form/) is a web page that shows
one of this package's instruments in the browser: the HiTOP-SR or one of its
modules, the HiTOP-BR, or the PID-5 in its full, short or brief form. The item
text, response options and instructions are this package's own, read from its
JSON export. The page needs no survey platform and no account, and it scores
nothing. Scoring is this package's job.

A study link tells the page what to show and where each participant's answers
go. Three destinations are offered: a file saved on the participant's own
device, a web address that accepts one row per participant, or a table in a
Supabase project. This article walks the web-address route end to end with a
Google Sheet as the store: a small script bound to the sheet receives one row
per participant, you download the sheet as a CSV file, and
`read_form_responses()` reads that file for the scoring functions. The steps
are in the order you take them.

## 1. Deploy the sheet's script

A Google Sheet accepts rows from the page through a Google Apps Script web app
bound to it. The script's code, and the steps to deploy it, are in the
hitop-form README under
[Send responses to a Google Sheet](https://github.com/jmgirard/hitop-form#send-responses-to-a-google-sheet).
In outline: create a new sheet, open Extensions and then Apps Script, replace
the contents of `Code.gs` with the README's code, save, and deploy it as a web
app that runs as you and that anyone can access. Copy the web app URL, which
ends in `/exec`.

The script appends one row per request to a tab named `Responses`, taking the
first row's keys as the header. It writes every cell as text, so a participant
code such as `007` keeps its zeros and a value such as `=1+1` stays those four
characters rather than becoming a formula. It refuses a request whose body is
not one object, holds more than 1000 keys, or has a key outside lower-case
letters, digits and underscores, so no one can grow the header without limit
or put a formula in it.

Deploy one sheet and script per form. A key the header lacks is added to it,
so a sheet that receives two forms' rows holds the union of their columns.
When the two forms are two instruments, `read_form_responses()` refuses the
download, because its item columns carry two stems (`hitopbr_01` beside
`pid5bf_01`). When they are a HiTOP-SR module and the full HiTOP-SR, the
columns share one stem, so the download reads as one data frame with the
module's rows blank on the items outside it, unless the module's link asked
for a random order: its rows' `item_order` then lists only the module's
items, and `read_form_responses()` refuses the download. Nothing else in the
download or in `read_form_responses()` tells the two apart.

## 2. Make the study link

Open the page's [link builder](https://jmgirard.github.io/hitop-form/link.html).
Choose the instrument (and, for a HiTOP-SR module, paste the descriptor
written by `write_module()`), and name the study. A link builder opened
from a "Make a study link" button on an instrument's download page arrives
with that instrument chosen, and one opened with a `c` parameter that
carries a module arrives with the descriptor pasted in as well. Give a
participant identifier if the
link is for one person; leave it empty for a link shared with many, and the
page asks each participant for one before the form starts. Under "Send
responses to", choose "A web address" and paste the `/exec` URL from step 1.
Press "Make the link" and send the link to the participants.

When a participant presses Finish, the page posts their answers to the
address as one row and waits for the script to confirm. On a confirmed send
the page says the responses were sent to the study team, and saves nothing.
When the send is not confirmed, because of an error, a lost connection, or no
answer within the page's limit, the page saves a CSV file on the
participant's device and names it, so the participant can send it to you by
hand. That screen also has a "Save the file" button, which saves the same
file again; its instructions tell the participant to press it if the file
did not appear. A row that reached the sheet after the limit is stored there
as well; the `submitted` column identifies such a pair.

## 3. Download the sheet as CSV

When collection ends, open the sheet's `Responses` tab and choose File, then
Download, then Comma Separated Values (.csv). The file has one header row and
one row per participant, in the page's column order: `study`, `participant`,
`instrument`, `form_build`, `submitted`, then one column per item. A link
that asks for a random order adds `item_order` after `submitted`, and its
item columns are in the instrument's order rather than the order shown. A
link that recruits through Prolific adds `prolific_study` and
`prolific_session` after `submitted`, or after `item_order` when the file
has it, before the item columns.
The script builds the header from the first row it receives and appends a
key it has not seen, so a sheet that already holds rows without `item_order`
puts that column after the item columns. `read_form_responses()` reads it
there. Start a new sheet for such a link if you want the column sixth.

Anyone who has the link can post a row to the sheet, because the script's URL
is inside the link. Look over the sheet before scoring it.

## 4. Read the file

`read_form_responses()` reads the download. The file below is one such
download, from a sheet that received two HiTOP-BR walks of the page. The
package installs it as an example, and `system.file()` gives its path.

```{r}
library(hitop)
path <- system.file("examples", "responses-sheet-hitopbr.csv", package = "hitop")
responses <- read_form_responses(path)
responses
```

Each row of the file is a row of the result. The lead columns are typed:
`participant` is character, so the codes `=1+1` and `007` come back as the
sheet stored them; `form_build` is a date; `submitted` is a UTC date-time.
Every item column is an integer. The sixth column, `item_order`, is the order
the participant saw the items, as item numbers, when the file records one and
`NA` when it does not, and scoring does not read it. The page records it
when the study link asks for a random order, which the link builder's
"Show the items in a random order" box sets. The seventh and eighth
columns, `prolific_study` and `prolific_session`, are the study and session
identifiers Prolific passed to the page when the link recruits through
Prolific, as [the Prolific route](#the-prolific-route) below describes,
and `NA` when it does not, as here. Scoring does not read them either.

A folder can hold the download beside any files participants sent by hand:
`read_form_responses()` reads every `.csv` file in it, file by file in path
order, and each file's rows in its own order. Every file must hold the same
item columns in the same order, so read each form's files in a call of their
own.

A file can also hold answer columns, which hold the answers to questions of
your own, each named `q_` and then the question's name, such as `q_age`.
The page writes them for a link with questions, as the section
[Your own questions](#your-own-questions) describes.
`read_form_responses()` accepts them anywhere after `submitted` and returns
them after the item columns. They come in the order the reader first meets
them, with the files in path order and each file read from left to right.
Each answer column is character, as written, and a blank cell is `NA`. Files
that hold different answer columns read together, and a row from a file
without a column holds `NA` in it. The answer columns do not enter the check
that every file holds the same item columns. After `q_`, a name must hold a
lower-case letter and then up to 29 lower-case letters, digits or
underscores. The reader stops with an error that names the file and the
column when a name does not. The reader does not convert an answer, so turn
a number answer into a number yourself:

```{r}
# The download above with an answer column added, as a sheet would hold it.
lines <- readLines(path, warn = FALSE)
lines[1] <- paste0(lines[1], ",q_age")
lines[-1] <- paste0(lines[-1], c(",34", ","))
answers_file <- tempfile(fileext = ".csv")
writeLines(lines, answers_file)

with_answers <- read_form_responses(answers_file)
with_answers$q_age
as.integer(with_answers$q_age)
```

## 5. Score it

Pass the item columns to the scoring function for the instrument. For the
HiTOP-BR that is `score_hitopbr()`:

```{r}
items <- grep("^hitopbr_", names(responses), value = TRUE)
score_hitopbr(responses, items = items, append = FALSE)
```

A full HiTOP-SR scores with `score_hitopsr()` the same way, and a PID-5 form
with `score_pid5()` and the matching `version`, as `vignette("pid5_scoring")`
shows. A HiTOP-SR module scores through the descriptor the page showed, with
`layout = "printed"`; the
[Building HiTOP-SR Modules](modules-hitopsr.html) article walks that step.

## The Supabase route

The page can insert each row into a table in a Supabase project instead. The
link builder shows the SQL that creates the table, and the hitop-form README
under
[Send responses to Supabase](https://github.com/jmgirard/hitop-form#send-responses-to-supabase)
walks the setup. The table's CSV export from the dashboard's Table Editor
reads with `read_form_responses()` exactly as the sheet's download does.

## The Prolific route

A study recruited through [Prolific](https://www.prolific.com/) gives each
participant a study URL. Prolific fills the participant's ID, the study's ID
and the submission's ID into that URL through three placeholders,
`{{%PROLIFIC_PID%}}`, `{{%STUDY_ID%}}` and `{{%SESSION_ID%}}`, as
[Prolific's API reference](https://docs.prolific.com/api-reference/studies/the-study-object)
states under the study's `external_study_url` field, and it asks that all
three be saved with the data
([What survey / experimental software is compatible with Prolific?](https://researcher-help.prolific.com/en/articles/445178-what-survey-experimental-software-is-compatible-with-prolific)).
The page fits that in two steps at the link builder.

Choose Prolific in the "Recruiting site" menu and leave the participant
field empty.
The builder then prints the link with the three placeholders on its end.
Paste that link as the study URL on Prolific's study page. When a
participant opens it, the page takes their Prolific ID as the participant
identifier and asks for none, and it writes the study and submission IDs
into `prolific_study` and `prolific_session`, after `submitted` (and after
`item_order` under the random order). A Supabase table for such a link
takes the two columns too, and the builder's SQL adds them. When you preview
the study on Prolific, the preview passes a 24-character ID in place of the
placeholder
([Previewing your study](https://researcher-help.prolific.com/en/articles/445131-previewing-your-study)),
so the preview walks the form as a participant would; a link opened with the
placeholders still in it, as the builder's "Open the link" does, asks for the
identifier instead. Prolific's own "I'll use URL parameters" option, in the
Data collection section of the study's setup, appends the three parameters
to the study URL itself
([the compatibility article](https://researcher-help.prolific.com/en/articles/445178-what-survey-experimental-software-is-compatible-with-prolific),
under "Recording participants' Prolific IDs"). A pasted link with the
builder's placeholders therefore carries each parameter once unless that
option adds it again; then the address holds the parameter twice, and the
page reads the first value that is neither blank nor a placeholder.

Give the completion URL from the study's page on Prolific, of the form
`https://app.prolific.com/submissions/complete?cc=…`
([the compatibility article](https://researcher-help.prolific.com/en/articles/445178-what-survey-experimental-software-is-compatible-with-prolific),
under "Returning Participants to Prolific"), in the builder's
"Completion URL" field. After a confirmed send the page shows the sent
screen, with a link to that address in place of "You can close this
page.", and then sends the participant there, which is the return Prolific
recommends
([Data collection](https://researcher-help.prolific.com/en/articles/445127-data-collection)).
After a saved file, because the link names no store or the send was not
confirmed, the page shows the file name first, then the instructions and
the "Save the file" button, and then a link to that address, so the
participant sees the file's name and can save it again before leaving.

A study can hold more than one completion code, each for its own outcome
([Custom completion codes](https://researcher-help.prolific.com/en/articles/445170-custom-completion-codes)),
and each code has its own `?cc=` address, as
[the API reference](https://docs.prolific.com/api-reference/studies/the-study-object)
states under `completion_codes`. To give a saved file its own code, put
that code's address in the builder's "Completion URL after a saved file"
field, beside the completion URL. The saved-file screens then link to it
in place of the completion URL, and a confirmed send still goes to the
completion URL.

The download then reads as above, with `prolific_study` and
`prolific_session` filled. The file below is one the page saved under such
a link, with the Prolific ID as the participant:

```{r}
path <- system.file("examples", "responses-hitopbr-prolific.csv", package = "hitop")
responses <- read_form_responses(path)
responses[, c("participant", "prolific_study", "prolific_session")]
```

## The SONA and CloudResearch Connect routes

Other recruiting sites can also pass each participant's identifier to the
page in the study link's address, under a parameter of their own. Choose
the site in the link builder's "Recruiting site" menu and leave the
participant field empty. The link then names the parameter in its `participantParam` field,
and the page takes the participant identifier from it. When the address
carries no value there, a blank one, or a placeholder the site did not fill,
the page asks for the identifier on its start screen. No column is added:
the identifier is the `participant` column, so the download reads as in the
steps above.

[SONA](https://www.sona-systems.com/) replaces the text `%SURVEY_CODE%` in
a study's Study URL with a number unique to the participant
([Using the SURVEY CODE Feature](https://www.sona-systems.com/researcher/using-the-survey-code-feature/)),
and its Qualtrics guide has the Study URL end in `?id=%SURVEY_CODE%`
([Qualtrics Help Page](https://www.sona-systems.com/help/qualtrics/)). With
SONA chosen, the builder names `id` as the parameter, and the printed link
ends in `&id=%SURVEY_CODE%`. Paste it as the Study URL in SONA. To grant
credit, SONA's client-side completion URL, of the form
`https://yourschool.sona-systems.com/webstudy_credit.aspx?experiment_id=123&credit_token=…&survey_code=XXXX`,
must be loaded with the participant's survey code in place of `XXXX`
([External Study Credit Granting](https://www.sona-systems.com/researcher/external-study-credit-granting/)).
Give that address in the builder's "Completion URL" field with
`{participant}` in place of `XXXX`. The page replaces `{participant}` with
the participant's identifier before it sends the participant there after a
confirmed send, and before it links there after a saved file (unless the
builder's "Completion URL after a saved file" names another address, which
takes `{participant}` the same way).

That completion URL carries a key specific to your study. SONA notes that a
participant can read it, because their browser loads the URL, and can use
it to try other survey codes and so grant credit to other participants
([Security Considerations](https://www.sona-systems.com/researcher/security-considerations/)).
Here the address sits inside the study link, so a participant can read the
key before they finish, not only at the end. SONA names its server-side
completion URL as the alternative. The external study loads that URL from
its own server, and the page has no server, so the page cannot use it.
A participant does not need to read the key to try another code. If the
address carries no survey code, for example because the participant removed
`id` from it, the start screen asks for the identifier. The page then puts
whatever the participant types into the completion URL.

[CloudResearch Connect](https://www.cloudresearch.com/) asks a study to
record each participant's Connect ID from a variable named `participantId`,
and to end either with a completion code or with the completion redirect
URL Connect gives the study
([How to Integrate your Survey with Connect](https://connect-researcher-help.cloudresearch.com/hc/en-us/articles/21181529476500-How-to-Integrate-your-Survey-with-Connect)).
With Connect chosen, the builder names `participantId` as the parameter and
adds nothing to the link. Give the link as the project URL in Connect, and
the completion redirect URL as the completion URL. Connect's `assignmentId`
and `projectId`, which that page says need not be collected, are not
recorded.

For another site, choose "Another site" and type the name of the parameter
that site adds to the address. The builder refuses `id` and
`participantId` there, because the SONA and Connect choices write those
names.

## Consent text before the form

The consent text belongs to you and your review board. The page shows the
text you give and adds no consent wording of its own. Paste the approved
text into the link builder's "Consent text" box. The page then shows the
text on a screen of its own, headed "Consent to take part", before the form
and its start screen. Under the text are an "I agree" and an "I do not
agree" button, and "I agree" goes on to the start screen.

The page reads the text as plain text. A blank line starts a new paragraph,
and a single line break stays a line break. A tag or an entity, such as
`<b>` or `&amp;`, shows as typed, so the text cannot carry formatting or
links. The text can hold up to 20,000 characters, and the builder refuses a
longer one.

"I do not agree" shows a closing screen with no way back to the form. The
page sends no answer and saves no file, so no row reaches the store and
`read_form_responses()` never sees the participant. The closing screen
shows the builder's "Declined text". When that box is empty, it shows "You
chose not to take part. You can close this page.", or only its first
sentence when the link sends the participant on, as the next sentences
describe. A recruiting site can give a study a completion address for this
outcome, such as a Prolific completion code for participants who do not
consent
([Custom completion codes](https://researcher-help.prolific.com/en/articles/445170-custom-completion-codes)).
Give it as the builder's "Completion URL after a decline". The link carries
it as its `completeDeclined` field, and the page sends a participant who
declines to that address.

A link with consent text travels as a `z` parameter in place of `c`, the
same fields compressed, so a long text makes a shorter link. The rows, the
file and the Supabase table have the same columns with consent text as
without it. The hitop-form README's section
[Show consent text before the form](https://github.com/jmgirard/hitop-form#show-consent-text-before-the-form)
has the details.

## Your own questions

A study link can carry questions of your own, such as age or how the
participant heard of the study. Add them in the link builder's "Your own
questions" part. Each question is asked either before the form or after
it. The page asks the before questions on a screen of its own after any
consent screen and ahead of the start screen. It asks the after questions
on a screen of its own after the last page of items. The questions never
share a screen with the items. A link holds up to 50 questions.

Each question has a name, and its answer is written in a column named `q_`
and then the name, after the item columns. `read_form_responses()` returns
these columns as the section on reading the file describes. A question has
one of four types, and each writes its answer as text:

- A text question writes the answer as typed.
- A whole-number question writes the number in digits, with no leading zero
  and a minus sign only below zero. A typed `007` is written `7`.
- A question with one choice writes the number of the chosen option,
  counted from 1 in the order the link lists the options.
- A question with any number of choices writes the numbers of the chosen
  options in ascending order, separated by single spaces, such as `1 3`.

An unanswered question writes an empty cell, which the reader returns as
`NA`. The numbers stand for positions in the link's list of options, so a
change to the order of the options between two links changes what the
numbers mean. Keep the order the same in every link of a study. To turn a
column of several choices into numbers, split it at the spaces:

```{r}
days <- c("1 3", "2", NA)
lapply(strsplit(days, " "), as.integer)
```

With a Supabase store, the builder's SQL adds one text column per question.
A table made before the questions were added has no column for them, so
Supabase refuses the row, and the page then saves the file on the
participant's device. The columns come from the question names alone, so
make a new table from the builder's SQL when you add a question or change a
name. A link with questions travels as a `z` parameter, as a
link with consent text does. The hitop-form README's section
[Ask your own questions](https://github.com/jmgirard/hitop-form#ask-your-own-questions)
has the details.

## Who holds the data

The answers go to the store the link names, and to a file on the
participant's device when the link names no store or the send is not
confirmed. They go nowhere else. With a Google Sheet, Google holds the rows;
with a Supabase table, Supabase holds them. The page's host, GitHub Pages, receives the study link with each page
load, and the link carries the instrument and any module, the study name, the
participant code, any consent text, the store's address and, for Supabase,
its key. It receives
no answers.

A participant code is whatever the link or the participant supplies. A code
you put in the link reaches the page's host with the link, and a code the
participant types is whatever they choose to type. If codes must not identify
anyone, issue pseudonymous codes in the links you send and keep the key to
them yourself. Under the Prolific route the code is the participant's
Prolific ID, which identifies them to Prolific, and the page's host receives
the three Prolific parameters, the ID among them, with each page load, as it
receives the rest of the address. Under the SONA and Connect routes the code
is the site's own identifier, and the host receives it the same way.

Whether a store may hold your study's data is a question for your
institution, and any agreement it needs with the store's vendor, such as a
business associate agreement or a data processing agreement, is the
institution's to make. This article describes where the rows go; it does not
say which store is suitable for which study.
