---
title: "Collecting Responses Online with hitop-form"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Collecting Responses Online with hitop-form}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[hitop-form](https://jmgirard.github.io/hitop-form/) is a web page that shows
one of this package's instruments in the browser: the HiTOP-SR or one of its
modules, the HiTOP-BR, or the PID-5 in its full, short or brief form. The item
text, response options and instructions are this package's own, read from its
JSON export. The page needs no survey platform and no account, and it scores
nothing. Scoring is this package's job.

A study link tells the page what to show and where each participant's answers
go. Three destinations are offered: a file saved on the participant's own
device, a web address that accepts one row per participant, or a table in a
Supabase project. This article walks the web-address route end to end with a
Google Sheet as the store: a small script bound to the sheet receives one row
per participant, you download the sheet as a CSV file, and
`read_form_responses()` reads that file for the scoring functions. The steps
are in the order you take them.

## 1. Deploy the sheet's script

A Google Sheet accepts rows from the page through a Google Apps Script web app
bound to it. The script's code, and the steps to deploy it, are in the
hitop-form README under
[Send responses to a Google Sheet](https://github.com/jmgirard/hitop-form#send-responses-to-a-google-sheet).
In outline: create a new sheet, open Extensions and then Apps Script, replace
the contents of `Code.gs` with the README's code, save, and deploy it as a web
app that runs as you and that anyone can access. Copy the web app URL, which
ends in `/exec`.

The script appends one row per request to a tab named `Responses`, taking the
first row's keys as the header. It writes every cell as text, so a participant
code such as `007` keeps its zeros and a value such as `=1+1` stays those four
characters rather than becoming a formula. It refuses a request whose body is
not one object, holds more than 1000 keys, or has a key outside lower-case
letters, digits and underscores, so no one can grow the header without limit
or put a formula in it.

Deploy one sheet and script per form. A key the header lacks is added to it,
so a sheet that receives two forms' rows, or a module's beside the full
instrument's, holds the union of their columns, and its download reads as one
data frame with the other form's items blank on every row. Nothing in the
download or in `read_form_responses()` tells the two apart.

## 2. Make the study link

Open the page's [link builder](https://jmgirard.github.io/hitop-form/link.html).
Choose the instrument (and, for a HiTOP-SR module, paste the descriptor
written by `write_module()`), and name the study. Give a participant identifier if the
link is for one person; leave it empty for a link shared with many, and the
page asks each participant for one before the form starts. Under "Send
responses to", choose "A web address" and paste the `/exec` URL from step 1.
Press "Make the link" and send the link to the participants.

When a participant presses Finish, the page posts their answers to the
address as one row and waits for the script to confirm. On a confirmed send
the page says the responses were sent to the study team, and saves nothing.
When the send is not confirmed, because of an error, a lost connection, or no
answer within the page's limit, the page saves a CSV file on the
participant's device and names it, so the participant can send it to you by
hand. A row that reached the sheet after the limit is stored there as well;
the `submitted` column identifies such a pair.

## 3. Download the sheet as CSV

When collection ends, open the sheet's `Responses` tab and choose File, then
Download, then Comma Separated Values (.csv). The file has one header row and
one row per participant, in the page's column order: `study`, `participant`,
`instrument`, `form_build`, `submitted`, then one column per item.

Anyone who has the link can post a row to the sheet, because the script's URL
is inside the link. Look over the sheet before scoring it.

## 4. Read the file

`read_form_responses()` reads the download. The file below is one such
download, from a sheet that received two HiTOP-BR walks of the page. The
package installs it as an example, and `system.file()` gives its path.

```{r}
library(hitop)
path <- system.file("examples", "responses-sheet-hitopbr.csv", package = "hitop")
responses <- read_form_responses(path)
responses
```

Each row of the file is a row of the result. The lead columns are typed:
`participant` is character, so the codes `=1+1` and `007` come back as the
sheet stored them; `form_build` is a date; `submitted` is a UTC date-time.
Every item column is an integer.

A folder can hold the download beside any files participants sent by hand:
`read_form_responses()` reads every `.csv` file in it, file by file in path
order, and each file's rows in its own order. Every file must hold the same
item columns in the same order, so read each form's files in a call of their
own.

## 5. Score it

Pass the item columns to the scoring function for the instrument. For the
HiTOP-BR that is `score_hitopbr()`:

```{r}
items <- grep("^hitopbr_", names(responses), value = TRUE)
score_hitopbr(responses, items = items, append = FALSE)
```

A full HiTOP-SR scores with `score_hitopsr()` the same way, and a PID-5 form
with `score_pid5()` and the matching `version`, as `vignette("pid5_scoring")`
shows. A HiTOP-SR module scores through the descriptor the page showed, with
`layout = "printed"`; the
[Building HiTOP-SR Modules](modules-hitopsr.html) article walks that step.

## The Supabase route

The page can insert each row into a table in a Supabase project instead. The
link builder shows the SQL that creates the table, and the hitop-form README
under
[Send responses to Supabase](https://github.com/jmgirard/hitop-form#send-responses-to-supabase)
walks the setup. The table's CSV export from the dashboard's Table Editor
reads with `read_form_responses()` exactly as the sheet's download does.

## Who holds the data

The answers go to the store the link names, and to a file on the
participant's device when the link names no store or the send is not
confirmed. They go nowhere else. With a Google Sheet, Google holds the rows;
with a Supabase table, Supabase holds them. The page's host, GitHub Pages, receives the study link with each page
load, and the link carries the instrument and any module, the study name, the
participant code, the store's address and, for Supabase, its key. It receives
no answers.

A participant code is whatever the link or the participant supplies. A code
you put in the link reaches the page's host with the link, and a code the
participant types is whatever they choose to type. If codes must not identify
anyone, issue pseudonymous codes in the links you send and keep the key to
them yourself.

Whether a store may hold your study's data is a question for your
institution, and any agreement it needs with the store's vendor, such as a
business associate agreement or a data processing agreement, is the
institution's to make. This article describes where the rows go; it does not
say which store is suitable for which study.
